Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2021-10-02 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

At very high $T_{leaf}$ (~40-60 $^{\circ}$C), photosystem II incurs irreversible damage, eventually leading to leaf necrosis and death [@bakerChlorophyllFluorescenceProbe2008; @feeley_thermal_2020; @kunertSummerTemperaturesReach]. 
Thermal tolerance is described in terms of the temperature at which efficiency of photosystem II starts to decrease, $T_{crit}$, and the $T_{leaf}$ at which efficiency of photosystem II had decreased by 50%, $T_{50}$ [@slotLeafHeatTolerance2021]. 
Thermal tolerances vary across species, with more variation explained by leaf traits than phylogeny [@feeley_thermal_2020; @slotLeafHeatTolerance2021]; for example, among tropical species, $T_{crit}$ and $T_{50}$ were found to be high in species with large leaves with high thermal capacitance and those with high LMA, respectively [@slotLeafHeatTolerance2021]. 
Thermal tolerances vary globally across latitude and climate, where $T_{50}$ and $T_{crit}$ decrease across increasing latitudes, and with elevation among tropical forests [@feeley_thermal_2020;  @slotLeafHeatTolerance2021; @osullivanThermalLimitsLeaf2017].
However, they are more closely adapted to microclimate than macroclimate [@feeley_thermal_2020; @slotLeafHeatTolerance2021]. 
For example, *Quercus muehlenbergii* growing in xeric, sunnier conditions showed higher $T_{50}$ by $2{^\circ}C$ than *Quercus macrocarpa* growing in shaded mesic conditions [@hamerlynckPhotosyntheticStomatalResponses1996a]. 
In Australia, along a vertical gradient within the canopy, *Acacia papyrocarpa* showed greater $T_{50}$ and higher composite climate stress in the lower, north-facing canopy than other canopy positions, correlating with low wind speed, greater radiation and $T_{air}$, and lower rates of heat dissipation, as indicated by longer leaf thermal time constants [@curtis_intracanopy_2019]. 
Across species sun leaves that experienced higher maximum temperatures show higher photosynthetic heat tolerance [@perezPhotosyntheticHeatTolerances2020]. 
However, considering sun and shade leaf differences within canopy (Table `r table_leaf_traits`), $T_{50}$ varies modestly, being slightly lower for shade than sun leaves for two of three tree species in Panama [@slotPhotosyntheticHeatTolerance2019]. 
The small difference in $T_{50}$ coupled with larger differences in $T_{leaf}$ across thermal microenvironments implies that more thermally tolerant leaves tend to operate closer to their thermal limits and could therefore be more vulnerable to heat anomalies [@perezPhotosyntheticHeatTolerances2020].

@@

Tree metabolism and growth are shaped by the positioning of their crowns within the vertical gradient.
Tree height, crown volume, and foliage biomass all scale with diameter at breast height (DBH), which in turn is a strong predictor of tree transpiration  [@meinzerWaterTransportTrees2001;@kunertRevisedHydrologicalModel2017; @anderson-teixeiraSizerelatedScalingTree2015] and photosynthesis.
Specifically, increases are linked to increased leaf area and the increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where higher light availability results in higher leaf area-specific photosynthesis (Table `r table_leaf_metabolism`).
The net foliar photosynthate production is allocated among functions including respiration, above-ground woody growth, foliar turnover, root growth and allocation to root-associated microorganisms, reproduction, defense, and storage of non-structural carbohydrates (NSCs). 
Among these, the process about which we know the most is woody aboveground growth, which consumes only a modest fraction of total photosynthate [~1/6 on the ecosystem level, @anderson-teixeiraCarbonCyclingMature2021] but is disproportionately important to long-term forest dynamics and carbon cycling in that it builds up woody tissues with a long residence time in the ecosystem [@russellResidenceTimesDecay2014].
Radial stem growth may increase or decrease over time as trees grow in DBH depending on the light environment [@anderson-teixeiraJointEffectsClimateinpress].
In open forests and for open-growth individual trees, growth rate declines with DBH [@muller-landauTestingMetabolicEcology2006; @anderson-teixeiraJointEffectsClimateinpress], whereas growth rate consistently increases with DBH for trees established in the understory of a closed-canopy forest [@muller-landauTestingMetabolicEcology2006; @anderson-teixeiraSizerelatedScalingTree2015]. 
This points to a dominant role of vertical profiles in microclimate in shaping tree growth rates within forests.

@@

In turn, the drought sensitivity of woody growth tends to be greater in canopy trees. 
In the field of dendrochronology, it is generally accepted that tree ring records of large, exposed trees are best suited for climate reconstructions because their annual growth displays the greatest sensitivity to interannual variation in climate [@frittsTreeRingsClimate1976]. 
However, only a relatively limited number of studies have directly examined drought- or temperature-sensitivities as a function of tree size.
These have most commonly found greater sensitivity to low precipitation or seasonally high temperatures among larger, more exposed trees [Fig. `r fig_treerings`, @trouillierSizeMattersComparison2018; @gillerotTreeNeighbourhoodDiversity2020; @mcgregorTreeHeightLeaf2021;@anderson-teixeiraJointEffectsClimateinpress], corroborating evidence from globally distributed forest censuses that larger trees tend to undergo larger growth declines during drought [@bennettLargerTreesSuffer2015]. 
In addition to lower drought resistance of growth, larger trees frequently exhibit lower ability to recover from stress [i.e., resilience, @mcgregorTreeHeightLeaf2021; @gillerotTreeNeighbourhoodDiversity2020], and greater increases in mortality [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019].
Mechanistically, this is almost certainly driven in part by the fact that larger trees have their crowns in a microenvironment that is more challenging during drought [Figs. `r fig_schematic` - `r fig_leaf_T`, @scharnweberConfessionsSolitaryOaks2019], yet there is also reason to believe that height itself provides disadvantages [@olsonPlantHeightHydraulic2018; @couvreurWaterTransportTall2018].  
Indeed, despite the potential for shorter trees in open forests to experience greater environmental stress [@curtis_intracanopy_2019], greater drought sensitivity of larger trees has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraJointEffectsClimateinpress], although there is also evidence that short trees in young stands are more drought-sensitive than taller trees in mature forests [@irvineAgerelatedChangesEcosystem2004].
The relative importance of exposure versus height in shaping drought sensitivity remains to be disentangled.

While it is clear that drought sensitivity increases with crown height though forest vertical gradients, and hence with DBH, it remains unclear how growth sensitivity to high $T_{air}$, independent of drought, varies across this gradient.
Because VPD increases rapidly with $T_{air}$, high $T_{air}$ is often associated with atmospheric drought, likely explaining negative growth responses -- particularly among larger trees -- even if soil moisture remains high [@novickIncreasingImportanceAtmospheric2016], or when precipitation is statistically accounted for [Fig. `r fig_treerings`, @anderson-teixeiraJointEffectsClimateinpress].
However, under conditions that are humid enough that canopy trees can maintain high $g_s$, we do not necessarily expect their photosynthesis (see section 4.2) or woody growth to exhibit higher sensitivity to $T_{air}$ than their understory counterparts. 
In eight mesic forests across the northeast United States, tree-ring records showed steep growth declines of understory trees at higher $T_{air}$, contrasting with the responses of canopy trees [Fig `r fig_treerings`, @rollinsonClimateSensitivityUnderstory2020].
Additional research is required to understand the mechanisms underlying these intriguing differences, and to disentangle size-related tree growth responses to hot-wet versus hot-dry conditions.

![**Figure `r fig_treerings`. Examples of tree-ring analyses showing differential interannual temperature sensitivity of annual growth between large canopy versus smaller understory trees.** In column (a), across three sites and species, trees with large diameter at breast height (DBH) had more negative growth responses to high temperatures during the current or previous growing season (denoted by c or p, respectively, on the x axes) than did small trees of the same species. Shown are tree ring width responses to the most influential temperature variable at the site ($T_{max}$ or PET) for the maximum and minimum tails of the DBH distribution included in the analysis.  Colored lines represent responses to the temperature variable in a model including a DBH - temperature interaction. Other model terms are held constant at their mean. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines and shading indicate the long-term mean $\pm$ 1 standard deviation of the temperature variable. From @anderson-teixeiraJointEffectsClimateinpress. In column (b), understory trees of four species had more negative growth responses to high growing season temperatures across eight New England forests. Growth is expressed as relativized basal area increment (BAI), where values > and < 100% indicate higher- or lower- than-average growth, respectively. Again, colored lines indicate modeled mean temperature responses and transparent ribbons indicate 95% confidence intervals. From @rollinsonClimateSensitivityUnderstory2020.](tree_rings/Figure_4_tree_rings.png)

@@

Canopy trees account for the majority of forest ecosystem water and carbon cycling.
While studies partitioning transpiration across forest vertical gradients are rare, both ET and transpiration have been shown to increase with height in a *Picea abies* forest in Germany, such that the upper half of the canopy contributed an estimated 80% of daytime ET [@staudtVerticalStructureEvapotranspiration2011]. 
Similarly, in a tropical forest in the Brazilian Amazon, canopy and subcanopy trees jointly contributed approximately 93% of ET, or 88% of transpiration [Fig `r fig_ecosystem`a; @kunertRevisedHydrologicalModel2017], and trees >33 cm DBH contributed >80% of transpiration in a forest in Costa Rica [@mooreUpscalingTranspirationDiverse2018]. 
In terms of C cycling, it has been estimated that canopy strata contribute $\ge$ 64% of net daytime CO~2~ uptake [i.e., GPP - ecosystem respiration, including from soil; @missonPartitioningForestCarbon2007]. 
Large trees also dominate in terms of woody above ground net primary productivity ($ANPP_{woody}$) and mortality ($M_{woody}$), with trees $\geq$ 10cm DBH usually contributing > 85% of $ANPP_{woody}$ and $M_{woody}$ across 25 globally distributed forests [Fig. `r fig_ecosystem`b, @piponiot_sizerelated_inreview].

![**Figure `r fig_ecosystem`. Vertical partitioning of (a) evapotranspiration and (b) C fluxes in tropical forests.** Panel (a) is from @kunertRevisedHydrologicalModel2017; panel (b) presents averages for 14 tropical forests from @piponiot_sizerelated_inreview.](schematics/ecosystem vertical partitioning/ecosystem partitioning.png)

@@

To the extent that warming is coupled to drought, we expect that the largest trees will usually be hardest-hit, particularly in cases where the drought is severe, and that this will occur at least in part because their crowns are positioned in a more challenging microenvironment [section 5.1, Figs. `r fig_schematic`- `r fig_treerings`, e.g., @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019; @anderson-teixeiraJointEffectsClimateinpress].
Specifically, warming will disproportionately stress tall canopy trees when $T_{air}$ and accompanying VPD rise enough that sun-exposed leaves cannot maintain both hydraulic safety and the transpirational cooling required to keep $T_{leaf}$ below damaging thresholds.
In contrast, understory trees will be more sheltered during droughts and heat waves, and in some settings may benefit from increased light availability [@bennettLargerTreesSuffer2015].
An interesting open question is to what extent these patterns vary with the nature of the drought, including the relative contributions to stress from low soil moisture versus high VPD. 
While the two are coupled over longer time scales [@humphreySoilMoistureAtmosphere2021; @novickIncreasingImportanceAtmospheric2016], the latter can be intense for short periods even when soil moisture is high (e.g., during a heat wave) and exerts a stronger influence on ET in many biomes [@novickIncreasingImportanceAtmospheric2016]. 
Given the mechanisms reviewed here, we might expect that atmospheric dryness in particular skews the stress more towards the exposed canopy trees.

What remains poorly understood is how responses to warming will vary across vertical gradients under mesic conditions. 
The leaf-level responses are relatively well-understood: at the metabolic level, photosynthesis and respiration are coupled to photosynthetic thermal optima, which reflect adapation to the local climate but do not vary consistently with height in the canopy (see section 4.2, Table 2). 
Beyond this inflection point, photosynthesis declines whereas respiration continues to increase exponentially, eventually shifting the carbon balance from sink to source, independent of water and irradiance [@duffyHowCloseAre2021]. 
While leaves display substantial plasticity to adapt to warmer temperatures  [@slotGeneralPatternsAcclimation2015; @slotPhotosyntheticPlasticityTropical2021; @wayJustRightTemperature2019;  @cunninghamTemperateRainforestTrees2003] and to recover from canopy heat stress [@smithEmpiricalEvidenceResilience2020], failure to fully acclimate will result in reduced carbon sequestration at leaf and ecosystem levels [@huangAirTemperatureOptima2019; @wayJustRightTemperature2019; @tanOptimumAirTemperature2017; @bennettThermalOptimaGross2021]. 
Exposed canopies are likely most vulnerable, as growing season $T_{air}$ is increasingly equaling or exceeding photosynthetic optima [@slotSituTemperatureResponse2017; @mauTemperateTropicalForest2018; @huangAirTemperatureOptima2019; @kumarathungeAcclimationAdaptationComponents2019]. 
Particularly in mid-latitude forests, which have the narrowest thermal safety margin between historical maximum temperatures and leaf thermal tolerance levels [@osullivanThermalLimitsLeaf2017], $T_{leaf}$ of sun-exposed leaves may approach thermal tolerance thresholds during heat waves, resulting in photosynthetic decline or even leaf death [e.g., @osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021; @kunertSummerTemperaturesReach].

>>>>

At very high $T_{leaf}$ (~40-60 $^{\circ}$C), photosystem II incurs irreversible damage, eventually leading to leaf necrosis and death [@bakerChlorophyllFluorescenceProbe2008; @feeley_thermal_2020; @kunertSummerTemperaturesReachinpress in press]. 
Thermal tolerance is described in terms of the temperature at which efficiency of photosystem II starts to decrease, $T_{crit}$, and the $T_{leaf}$ at which efficiency of photosystem II had decreased by 50%, $T_{50}$ [@slotLeafHeatTolerance2021]. 
Thermal tolerances vary across species, with more variation explained by leaf traits than phylogeny [@feeley_thermal_2020; @slotLeafHeatTolerance2021]; for example, among tropical species, $T_{crit}$ and $T_{50}$ were found to be high in species with large leaves with high thermal capacitance and those with high LMA, respectively [@slotLeafHeatTolerance2021]. 
Thermal tolerances vary globally across latitude and climate, where $T_{50}$ and $T_{crit}$ decrease across increasing latitudes, and with elevation among tropical forests [@feeley_thermal_2020;  @slotLeafHeatTolerance2021; @osullivanThermalLimitsLeaf2017].
However, they are more closely adapted to microclimate than macroclimate [@feeley_thermal_2020; @slotLeafHeatTolerance2021]. 
For example, *Quercus muehlenbergii* growing in xeric, sunnier conditions showed higher $T_{50}$ by $2{^\circ}C$ than *Quercus macrocarpa* growing in shaded mesic conditions [@hamerlynckPhotosyntheticStomatalResponses1996a]. 
In Australia, along a vertical gradient within the canopy, *Acacia papyrocarpa* showed greater $T_{50}$ and higher composite climate stress in the lower, north-facing canopy than other canopy positions, correlating with low wind speed, greater radiation and $T_{air}$, and lower rates of heat dissipation, as indicated by longer leaf thermal time constants [@curtis_intracanopy_2019]. 
Across species sun leaves that experienced higher maximum temperatures show higher photosynthetic heat tolerance [@perezPhotosyntheticHeatTolerances2020]. 
However, considering sun and shade leaf differences within canopy (Table `r table_leaf_traits`), $T_{50}$ varies modestly, being slightly lower for shade than sun leaves for two of three tree species in Panama [@slotPhotosyntheticHeatTolerance2019]. 
The small difference in $T_{50}$ coupled with larger differences in $T_{leaf}$ across thermal microenvironments implies that more thermally tolerant leaves tend to operate closer to their thermal limits and could therefore be more vulnerable to heat anomalies [@perezPhotosyntheticHeatTolerances2020].

@@

Tree metabolism and growth are shaped by the positioning of their crowns within the vertical gradient.
Tree height, crown volume, and foliage biomass all scale with diameter at breast height (DBH), which in turn is a strong predictor of tree transpiration  [@meinzerWaterTransportTrees2001;@kunertRevisedHydrologicalModel2017; @anderson-teixeiraSizerelatedScalingTree2015] and photosynthesis.
Specifically, increases are linked to increased leaf area and the increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where higher light availability results in higher leaf area-specific photosynthesis (Table `r table_leaf_metabolism`).
The net foliar photosynthate production is allocated among functions including respiration, above-ground woody growth, foliar turnover, root growth and allocation to root-associated microorganisms, reproduction, defense, and storage of non-structural carbohydrates (NSCs). 
Among these, the process about which we know the most is woody aboveground growth, which consumes only a modest fraction of total photosynthate [~1/6 on the ecosystem level, @anderson-teixeiraCarbonCyclingMature2021] but is disproportionately important to long-term forest dynamics and carbon cycling in that it builds up woody tissues with a long residence time in the ecosystem [@russellResidenceTimesDecay2014].
Radial stem growth may increase or decrease over time as trees grow in DBH depending on the light environment [@anderson-teixeiraJointEffectsClimateinpress in press].
In open forests and for open-growth individual trees, growth rate declines with DBH [@muller-landauTestingMetabolicEcology2006; @anderson-teixeiraJointEffectsClimateinpress in press], whereas growth rate consistently increases with DBH for trees established in the understory of a closed-canopy forest [@muller-landauTestingMetabolicEcology2006; @anderson-teixeiraSizerelatedScalingTree2015]. 
This points to a dominant role of vertical profiles in microclimate in shaping tree growth rates within forests.

@@

In turn, the drought sensitivity of woody growth tends to be greater in canopy trees. 
In the field of dendrochronology, it is generally accepted that tree ring records of large, exposed trees are best suited for climate reconstructions because their annual growth displays the greatest sensitivity to interannual variation in climate [@frittsTreeRingsClimate1976]. 
However, only a relatively limited number of studies have directly examined drought- or temperature-sensitivities as a function of tree size.
These have most commonly found greater sensitivity to low precipitation or seasonally high temperatures among larger, more exposed trees [Fig. `r fig_treerings`, @trouillierSizeMattersComparison2018; @gillerotTreeNeighbourhoodDiversity2020; @mcgregorTreeHeightLeaf2021;@anderson-teixeiraJointEffectsClimateinpress in press], corroborating evidence from globally distributed forest censuses that larger trees tend to undergo larger growth declines during drought [@bennettLargerTreesSuffer2015]. 
In addition to lower drought resistance of growth, larger trees frequently exhibit lower ability to recover from stress [i.e., resilience, @mcgregorTreeHeightLeaf2021; @gillerotTreeNeighbourhoodDiversity2020], and greater increases in mortality [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019].
Mechanistically, this is almost certainly driven in part by the fact that larger trees have their crowns in a microenvironment that is more challenging during drought [Figs. `r fig_schematic` - `r fig_leaf_T`, @scharnweberConfessionsSolitaryOaks2019], yet there is also reason to believe that height itself provides disadvantages [@olsonPlantHeightHydraulic2018; @couvreurWaterTransportTall2018].  
Indeed, despite the potential for shorter trees in open forests to experience greater environmental stress [@curtis_intracanopy_2019], greater drought sensitivity of larger trees has been observed in open as well as closed-canopy forests [@bennettLargerTreesSuffer2015; @anderson-teixeiraJointEffectsClimateinpress in press], although there is also evidence that short trees in young stands are more drought-sensitive than taller trees in mature forests [@irvineAgerelatedChangesEcosystem2004].
The relative importance of exposure versus height in shaping drought sensitivity remains to be disentangled.

While it is clear that drought sensitivity increases with crown height though forest vertical gradients, and hence with DBH, it remains unclear how growth sensitivity to high $T_{air}$, independent of drought, varies across this gradient.
Because VPD increases rapidly with $T_{air}$, high $T_{air}$ is often associated with atmospheric drought, likely explaining negative growth responses -- particularly among larger trees -- even if soil moisture remains high [@novickIncreasingImportanceAtmospheric2016], or when precipitation is statistically accounted for [Fig. `r fig_treerings`, @anderson-teixeiraJointEffectsClimateinpress in press].
However, under conditions that are humid enough that canopy trees can maintain high $g_s$, we do not necessarily expect their photosynthesis (see section 4.2) or woody growth to exhibit higher sensitivity to $T_{air}$ than their understory counterparts. 
In eight mesic forests across the northeast United States, tree-ring records showed steep growth declines of understory trees at higher $T_{air}$, contrasting with the responses of canopy trees [Fig `r fig_treerings`, @rollinsonClimateSensitivityUnderstory2020].
Additional research is required to understand the mechanisms underlying these intriguing differences, and to disentangle size-related tree growth responses to hot-wet versus hot-dry conditions.

![**Figure `r fig_treerings`. Examples of tree-ring analyses showing differential interannual temperature sensitivity of annual growth between large canopy versus smaller understory trees.** In column (a), across three sites and species, trees with large diameter at breast height (DBH) had more negative growth responses to high temperatures during the current or previous growing season (denoted by c or p, respectively, on the x axes) than did small trees of the same species. Shown are tree ring width responses to the most influential temperature variable at the site ($T_{max}$ or PET) for the maximum and minimum tails of the DBH distribution included in the analysis.  Colored lines represent responses to the temperature variable in a model including a DBH - temperature interaction. Other model terms are held constant at their mean. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines and shading indicate the long-term mean $\pm$ 1 standard deviation of the temperature variable. From @anderson-teixeiraJointEffectsClimateinpress in press. In column (b), understory trees of four species had more negative growth responses to high growing season temperatures across eight New England forests. Growth is expressed as relativized basal area increment (BAI), where values > and < 100% indicate higher- or lower- than-average growth, respectively. Again, colored lines indicate modeled mean temperature responses and transparent ribbons indicate 95% confidence intervals. From @rollinsonClimateSensitivityUnderstory2020.](tree_rings/Figure_4_tree_rings.png)

@@

Canopy trees account for the majority of forest ecosystem water and carbon cycling.
While studies partitioning transpiration across forest vertical gradients are rare, both ET and transpiration have been shown to increase with height in a *Picea abies* forest in Germany, such that the upper half of the canopy contributed an estimated 80% of daytime ET [@staudtVerticalStructureEvapotranspiration2011]. 
Similarly, in a tropical forest in the Brazilian Amazon, canopy and subcanopy trees jointly contributed approximately 93% of ET, or 88% of transpiration [Fig `r fig_ecosystem`a; @kunertRevisedHydrologicalModel2017], and trees >33 cm DBH contributed >80% of transpiration in a forest in Costa Rica [@mooreUpscalingTranspirationDiverse2018]. 
In terms of C cycling, it has been estimated that canopy strata contribute $\ge$ 64% of net daytime CO~2~ uptake [i.e., GPP - ecosystem respiration, including from soil; @missonPartitioningForestCarbon2007]. 
Large trees also dominate in terms of woody above ground net primary productivity ($ANPP_{woody}$) and mortality ($M_{woody}$), with trees $\geq$ 10cm DBH usually contributing > 85% of $ANPP_{woody}$ and $M_{woody}$ across 25 globally distributed forests [Fig. `r fig_ecosystem`b, @piponiotSizerelatedTreeBiomassinreview in review].

![**Figure `r fig_ecosystem`. Vertical partitioning of (a) evapotranspiration and (b) C fluxes in tropical forests.** Panel (a) is from @kunertRevisedHydrologicalModel2017; panel (b) presents averages for 14 tropical forests from @piponiotSizerelatedTreeBiomassinreview in review.](schematics/ecosystem vertical partitioning/ecosystem partitioning.png)

@@

To the extent that warming is coupled to drought, we expect that the largest trees will usually be hardest-hit, particularly in cases where the drought is severe, and that this will occur at least in part because their crowns are positioned in a more challenging microenvironment [section 5.1, Figs. `r fig_schematic`- `r fig_treerings`, e.g., @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019; @anderson-teixeiraJointEffectsClimateinpress in press].
Specifically, warming will disproportionately stress tall canopy trees when $T_{air}$ and accompanying VPD rise enough that sun-exposed leaves cannot maintain both hydraulic safety and the transpirational cooling required to keep $T_{leaf}$ below damaging thresholds.
In contrast, understory trees will be more sheltered during droughts and heat waves, and in some settings may benefit from increased light availability [@bennettLargerTreesSuffer2015].
An interesting open question is to what extent these patterns vary with the nature of the drought, including the relative contributions to stress from low soil moisture versus high VPD. 
While the two are coupled over longer time scales [@humphreySoilMoistureAtmosphere2021; @novickIncreasingImportanceAtmospheric2016], the latter can be intense for short periods even when soil moisture is high (e.g., during a heat wave) and exerts a stronger influence on ET in many biomes [@novickIncreasingImportanceAtmospheric2016]. 
Given the mechanisms reviewed here, we might expect that atmospheric dryness in particular skews the stress more towards the exposed canopy trees.

What remains poorly understood is how responses to warming will vary across vertical gradients under mesic conditions. 
The leaf-level responses are relatively well-understood: at the metabolic level, photosynthesis and respiration are coupled to photosynthetic thermal optima, which reflect adapation to the local climate but do not vary consistently with height in the canopy (see section 4.2, Table 2). 
Beyond this inflection point, photosynthesis declines whereas respiration continues to increase exponentially, eventually shifting the carbon balance from sink to source, independent of water and irradiance [@duffyHowCloseAre2021]. 
While leaves display substantial plasticity to adapt to warmer temperatures  [@slotGeneralPatternsAcclimation2015; @slotPhotosyntheticPlasticityTropical2021; @wayJustRightTemperature2019;  @cunninghamTemperateRainforestTrees2003] and to recover from canopy heat stress [@smithEmpiricalEvidenceResilience2020], failure to fully acclimate will result in reduced carbon sequestration at leaf and ecosystem levels [@huangAirTemperatureOptima2019; @wayJustRightTemperature2019; @tanOptimumAirTemperature2017; @bennettThermalOptimaGross2021]. 
Exposed canopies are likely most vulnerable, as growing season $T_{air}$ is increasingly equaling or exceeding photosynthetic optima [@slotSituTemperatureResponse2017; @mauTemperateTropicalForest2018; @huangAirTemperatureOptima2019; @kumarathungeAcclimationAdaptationComponents2019]. 
Particularly in mid-latitude forests, which have the narrowest thermal safety margin between historical maximum temperatures and leaf thermal tolerance levels [@osullivanThermalLimitsLeaf2017], $T_{leaf}$ of sun-exposed leaves may approach thermal tolerance thresholds during heat waves, resulting in photosynthetic decline or even leaf death [e.g., @osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021; @kunertSummerTemperaturesReachinpress in press].
